Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-02-28 application.tex
redoing app section

\section{Application to protoplanetary disks}\label{application} 
As an application of our results, we estimate where in a 
protoplanetary disk do thermodynamic conditions allow the VSI to 
operate. Specifically, we consider the fundamental VSI, 
for which we can use the criterion Eq. \ref{iso_vsi_cond}.

%In our disk models we have adopted a simple thermal relaxation 
%function in the energy equation. 
For a realistic disk one can consider the energy equation with
radiative cooling, 
%To connect our model with a realistic
%disk, we consider a more physically-motivated form of the energy
%equation with radiative cooling 
\begin{align}\label{real_energy}
%\rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
  c_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}where the temperature $T$ is defined through $P=\mathcal{R}\rho
T/\mu$, $\mathcal{R}$ is the ideal gas constant and $\mu$ is
the mean molecular weight, $c_v$ is the heat capacity at constant
volume, 
% In Eq. \ref{real_energy} $S$ is the
% specific entropy,
% \begin{align}
%   S \equiv c_v\ln{\left(\frac{p}{\rho^{\gamma}}\right)} 
% \end{align}
% where $c_v \equiv \left(\gamma-1\right)^{-1}\mathcal{R}/\mu$ is the
% specific heat capcity at constant volume
$\bm{F}$ is the radiative flux, and $Q_+$ is a heating
term such that an equilibrium solution exists.

One should now model in detail each of the terms on the right
hand side (RHS) of Eq. \ref{real_energy}, linearize them, then solve
the resulting linear eigenvalue problem. Such an approach is beyond
the scope of this paper. Here, we simply model the RHS of
Eq. \ref{real_energy} such that it can be represented by a constant
thermal relaxation time $t_c$ as used in our linear calculations.  
%Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
%the 
To do so, we first consider separately perturbations
with radial lengthscales $l\equiv 2\pi/k_x\gg l_\mathrm{rad}$ and 
$l\ll l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity.

@@

\subsection{A model for combined thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
 % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
%where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
%$\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
%and Eq. \ref{eta_def}. 
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$.

In order to utilize our linear results, which assumes $t_c$ is a
constant, we need to make further 
approximations. Specifically, we consider vertically-isothermal disks
and replace explicit $\rho$ dependencies in the above expressions by $\Sigma/
H$, where $\Sigma$ is the surface density at the fiducial
radius of interest. For example, the dimensionless diffusion
coefficient becomes 
\begin{align}\label{diff_coeff_dimensionless}
  \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
    c_v\Omega_k}. 
\end{align}
Furthermore, we will choose a temperature-dependent opacity model for
simplicity, so that $\kappa_d=\kappa_d(T)$. The dimensionless thermal
relaxation time $\beta$  
\begin{align}\label{real_beta}
  \beta \equiv t_c\Omega_k =
  \frac{1}{\hat{\eta}}\left(\frac{l_\mathrm{rad}^2}{3H^2}
    + \frac{1}{\khat^2}\right)  \simeq
  \frac{1}{\hat{\eta}}\left(\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{1}{\khat^2}\right)
\end{align}
is then a constant for vertically-isothermal disks, 
and we can directly apply previous results.

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $r_\mathrm{AU}\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
\beta =
\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)r_\mathrm{AU}^{-57/14}\left(\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}r_\mathrm{AU}^{33/7}+\frac{1}{\khat^2}\right).    
\end{align}
We refer to the case with the scalings
$\hat{\Sigma}=\hat{\kappa}_d=\hat{T}=1$ as the Minimum Mass Solar
Nebulae (MMSN).  We also evaluate the criterion Eq. \ref{iso_vsi_cond} for the fiducial
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is necessary for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
%becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn} and
Eq. \ref{bcrit_mmsn} with with $\mu =2.33$ and $\gamma=1.4$.  We set
the scaling $\hat{T}=\hat{\Sigma}=1$ and consider three opacity scales:
$\hat{\kappa}_d=0.1, \,1 $ and $10$.

The figure suggests that the fundamental VSI 
operates in the outer parts of the MMSN at radii on 
the order of tens of AU ($\hat{\kappa}_d=1$, middle panel). 
However, the thermal relaxation time $\beta$
is not much shorter than $\beta_\mathrm{crit}$, so the VSI may be 
weak. Lowering the opacity shifts the curves
inward, but raises also them. In fact, lowering the opacity by an
order of magnitude renders the fiducial  disk stable to the fundametal
mode for the range of wavenumbers considered ($\hat{\kappa}_d=0.1$,
upper panel). Raising the opacity, on
the other hand, decreases the thermal timescales, but also shifts the
curves outward, so the fundamental mode is restricted to the outer
disk ($\hat{\kappa}_d=10$, lower panel).

@@

\begin{figure}
  \includegraphics[scale=.43,clip=true,trim=0cm 1.8cm 0cm 0cm]{figures/bcrit_mmsn0d1} 
  \includegraphics[scale=.43,clip=true,trim=0cm 1.8cm 0cm 1cm]{figures/bcrit_mmsn}  
  \includegraphics[scale=.43,clip=true,trim=0cm 0cm 0cm 1cm]{figures/bcrit_mmsn10} 
  \caption{Dimensionless thermal relaxation timescales $\beta$ in a
    fiducial protoplanetary disk model. Eq. \ref{beta_mmsn} is plotted
    for three values of the 
    perturbation radial wavenumber: $\khat=1$ (dotted), $\khat=10$
    (dashed) and $\khat=100$ (dashed-dot), for three values of the
    opacity relative to the MMSN: $\hat{\kappa}_d=0.1$ (top),
    $\hat{\kappa}_d=1$ (middle) and $\hat{\kappa}_d=10$ (bottom).  
    The solid line is the
    critical thermal relaxation timescale $\beta_\mathrm{crit}$. 
    % and 
    % $\beta<\beta_\mathrm{crit}$ is necessary for the fundamental isothermal VSI to operate  
    % for $\khat\gg1$. 
    \label{mmsn_bcrit_bcool}}   
\end{figure}

@@

\subsection{Numerical example}\label{mmsn_example}
Consider  the MMSN at $r=50\mathrm{AU}$. Eq. \ref{mmsn_epsilon} gives
$\epsilon \simeq 
0.067$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.072. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=30$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.049.
\end{align*}
(In fact, Fig. \ref{mmsn_bcrit_bcool} shows that $\beta$  is not
sensitive to $\khat$ at $r=50\mathrm{AU}$ for $\khat\gg1$.)
The MMSN is unstable to the fundamental VSI, but not
strongly.

In Fig. \ref{mmsn_growth}, we calculate the fundamental
VSI growth rates using the MMSN disk parameters at $r=50\mathrm{AU}$
and the scale-dependent thermal relaxation $\beta(\khat)$ given by
Eq. \ref{beta_mmsn} in our linear code.  In this case we found a 
larger vertical domain with $\zmax=7\epsilon r$ was needed to capture
modes with $\khat\gtrsim60$, which have perturbations
concentrated near the vertical boundaries. However, modes with
$\khat\lesssim 60$ were found to be insensitive to boundary
conditions.

The most unstable mode occurs at $\khat\simeq30$ with growth rate 
\begin{align*}
  \nu =0.058\epsilon\Omega_k = 3.9\times10^{-3}\Omega_k.  
\end{align*}
This corresponds to a characteristic growth time $\sim
40$ orbits. However, since growth rates are weakly dependent on $\khat$ for
$10\lesssim \khat \lesssim 60$, this may be taken as a representative growth
time of the fundamental VSI in the MMSN at $r=50\mathrm{AU}$. The
eigenfunction for $\khat=30$ is shown in Fig. \ref{mmsn_eigenW}.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/growth_mmsn_50AU} 
  \caption{VSI growth rates in the MMSN at
    $r=50\mathrm{AU}$. The thermal relaxation timescale $\beta(\khat)$
    used for this calculation is given by
    Eq. \ref{beta_mmsn} and is scale-dependent. 
    \label{mmsn_growth}}  
\end{figure}

% Fig. \ref{mmsn_eigenW} show several eigenfunctions for the modes in
% Fig. \ref{mmsn_growth}. As $\khat$ is increased, the disturbance
% become more concentrated near the vertical boundaries. Thus, modes
% with very large $\khat$ may not be important in practice because they
% predominantly affect the disk atmosphere, which contains little mass.

\begin{figure}
   \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorW_mmsnkx30.ps}
  % \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  % 0cm]{figures/eigenvectorW_mmsnkx10.ps} 
  % \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  % 0cm]{figures/eigenvectorW_mmsnkx50.ps} 
  \caption{Fundamental VSI mode in the MMSN at 
    $r=50\mathrm{AU}$. The eigenfunction $W$ for $\khat=30$ is shown. The real (imaginary)
    part of $W$ is plotted as the solid (dotted)
    lines. The thermal relaxation timescale $\beta(\khat)$ is given by
    Eq. \ref{beta_mmsn} and is scale-dependent. \label{mmsn_eigenW}}    
\end{figure}
% only find one unstable mode -> close to marginal 
% larger kx

>>>>

\section{Application to protoplanetary disks}\label{application} 
As a first application of our linear models, we estimate where in a
protoplanetary disk do the thermodynamic conditions allow the VSI to
operate. For a realistic disk one can consider the energy equation with
radiative cooling, 
%To connect our model with a realistic
%disk, we consider a more physically-motivated form of the energy
%equation with radiative cooling 
\begin{align}\label{real_energy}
%\rho T \frac{DS}{Dt} = - \left(\nabla\cdot\bm{F} - Q_+\right), 
\frac{\p P}{\p t} + \bm{v}\cdot\nabla P +\gamma P \nabla\cdot\bm{v} = - \frac{P}{\rho T
  c_v}\left(\nabla\cdot\bm{F} - Q_+\right), 
\end{align}where the temperature $T$ is defined through $P=\mathcal{R}\rho
T/\mu$, $\mathcal{R}$ is the ideal gas constant and $\mu$ is
the mean molecular weight, $c_v$ is the heat capacity at constant
volume, 
% In Eq. \ref{real_energy} $S$ is the
% specific entropy,
% \begin{align}
%   S \equiv c_v\ln{\left(\frac{p}{\rho^{\gamma}}\right)} 
% \end{align}
% where $c_v \equiv \left(\gamma-1\right)^{-1}\mathcal{R}/\mu$ is the
% specific heat capcity at constant volume
$\bm{F}$ is the radiative flux, and $Q_+$ is a heating
term such that an equilibrium solution exists.

One should now model in detail each of the terms on the right
hand side (RHS) of Eq. \ref{real_energy}, linearize them, then solve
the resulting linear eigenvalue problem. Such an approach is beyond
the scope of this paper. Here, we simply model the RHS of
Eq. \ref{real_energy} such that it can be represented by a 
thermal relaxation time $t_c$ for the disk temperature. 
%Our aim is to relate the right hand side (RHS) of Eq. \ref{real_energy} to
%the 
To do so, we first consider separately perturbations
with radial lengthscales $l\equiv 2\pi/k_x\gg l_\mathrm{rad}$ and 
$l\ll l_\mathrm{rad}$, where      
\begin{align}\label{lrad}
  l_\mathrm{rad} \equiv \frac{1}{\kappa_d\rho} 
\end{align} 
is the photon mean-free-path and $\kappa_d$ is the (dust) opacity.

@@

\subsection{A model for combined thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff}. %=
 % \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
%where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
%$\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
%and Eq. \ref{eta_def}. 
Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$. Writing the
volume density $\rho = \Sigma\hat{\rho}/\sqrt{2\pi}H$, where
$\hat{\rho} = \rho/\rho_0$ and $\Sigma$ is the surface density at the
radius of interest, the 
% In order to utilize our linear results, which assumes $t_c$ is a
% constant, we need to make further 
% approximations. Specifically, we consider vertically-isothermal disks
% and replace explicit $\rho$ dependencies in the above expressions by $\Sigma/
% H$, where $\Sigma$ is the surface density at the fiducial
% radius of interest. For example, the dimensionless diffusion
% coefficient becomes 
% \begin{align}\label{diff_coeff_dimensionless}
%   \hat{\eta} \to  \frac{16\sigma_s T^3}{3\kappa_d\Sigma^2
%     c_v\Omega_k}. 
% \end{align}
% Furthermore, we will choose a temperature-dependent opacity model for
% simplicity, so that $\kappa_d=\kappa_d(T)$.
dimensionless thermal
relaxation time $\beta$ becomes 
\begin{align}\label{real_beta}
  \beta(z;\khat) \equiv t_c\Omega_k =
  % \frac{1}{\hat{\eta}}\left[\frac{l_\mathrm{rad}^2}{3H^2}
  %   + \frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right]  \simeq
  \frac{1}{\hat{\eta}}\left[\frac{1}{3\kappa_d^2\Sigma^2} 
    + \frac{\hat{\rho}^2(z)}{2\pi \khat^2}\right].
\end{align}
Note that the thermal relaxation time depends on both the perturbation lengthscale $\khat$ 
and distance from the midplane. %The latter is due to the density-dependence in the diffusion time. 
%is then a constant for vertically-isothermal disks, 
%and we can directly apply previous results.

\subsection{Evaluation for a fiducial disk model}
We evaluate Eq. \ref{real_beta} for the protoplanetary disk model
described in \cite{chiang10} and summarized in Appendix \ref{mmsn}
(where an opacity model is also chosen). Defining $r_\mathrm{AU}\equiv
r/\mathrm{AU}$, we find Eq. \ref{real_beta} becomes
\begin{align}\label{beta_mmsn}
  \beta(z; \khat) =
  &\frac{4.4\times10^5}{\mu(\gamma-1)}\left(\frac{\hat{\kappa}_d\hat{\Sigma}^2}{\hat{T}}\right)r_\mathrm{AU}^{-57/14}\notag\\
&\times\left[\frac{8.3\times10^{-9}}{\hat{\kappa}_d^2\hat{T}^4\hat{\Sigma}^2}r_\mathrm{AU}^{33/7}+\frac{\hat{\rho}^2(z)}{2\pi\khat^2}\right].     
\end{align}
We refer to the case with the scalings
$\hat{\Sigma}=\hat{\kappa}_d=\hat{T}=1$ as the Minimum Mass Solar
Nebulae (MMSN). The physical thermal relaxation timescale
$\beta$ is height-dependent, whilst for analytical discussion we
assumed it is constant. Nevertheless, it is instructive to evaluate
and compare with the criterion Eq. \ref{iso_vsi_cond} for the fiducial  
disk model,  
\begin{align}\label{bcrit_mmsn}
  \beta_\mathrm{crit} = \frac{1.44\times10^{-2}}{(\gamma
    -1)}\left(\frac{\hat{T}}{\mu}\right)^{1/2}r_\mathrm{AU}^{2/7}. 
\end{align}
We recall $\beta < \beta_\mathrm{crit}$ is necessary for the
fundamental VSI when $\khat\gg 1$. %For $\khat\to0$, this
%becomes a sufficient condition (see Fig. \ref{relax_bcrit} and
%\S\ref{iso_vsi_beta_crit}).

In Fig. \ref{mmsn_bcrit_bcool} we plot Eq. \ref{beta_mmsn}  with $\mu
=2.33$ and $\gamma=1.4$.  We set the scaling $\hat{T}=\hat{\Sigma}=1$
and consider three opacity scales: $\hat{\kappa}_d=0.1, \,1 $ and $10$
at two heights: $z=0$ and $z=3H$. We also plot Eq. \ref{bcrit_mmsn} in
each panel for comparison.

Consider first the MMSN shown in the middle panels of
Fig. \ref{mmsn_bcrit_bcool}. The critical thermal timescale,
Eq. \ref{bcrit_mmsn}, as derived from our analytical discussion, is
most useful to assess the VSI in the optically thin regime (where the
curves overlap), since in this limit $\beta$ does not depend on
height. We can therefore conclude that the VSI can operate at tens of
AU in the MMSN.

In the inner few AU of the MMSN, however, $\beta$  has a
strong dependence on height because the disk is optically thick. Given
a fixed $\khat$, one can have $\beta > \beta_\mathrm{crit}$ at the
midplane, but eventually $\beta<\beta_\mathrm{crit}$ with increasing
height. Our analytical criteria is not valid for variable $\beta$, but
given that the VSI depends on vertical shear, an effect that becomes
important away from the disk midplane, the inner disk may still be
able to develop the VSI, but it is likely restricted to the disk
surface.

Now consider lowering the opacity, as shown in the top panels of
Fig. \ref{mmsn_bcrit_bcool}. The curves shift inward, but raises also
them. In fact, lowering the opacity by an order of magnitude renders
$\beta > \beta_\mathrm{crit}$ even away from the midplane. This
suggests that fiducial  disk stable to the fundametal mode for the
range of wavenumbers considered.

Raising the opacity, as shown in the lower panels of
Fig. \ref{mmsn_bcrit_bcool}, decreases the thermal timescales, but
also shifts the curves outward, so the fundamental mode is restricted
to larger radii.

@@

\begin{figure*}
  \includegraphics[scale=.47,clip=true,trim=0cm 1.8cm 0cm
  0cm]{figures/bcrit_mmsn_kap0d1_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm 1.8cm 0cm
  0cm]{figures/bcrit_mmsn_kap0d1_z3}\\ 
  \includegraphics[scale=.47,clip=true,trim=0cm 1.8cm 0cm
  1cm]{figures/bcrit_mmsn_kap1_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm
  1.8cm 0cm 1cm]{figures/bcrit_mmsn_kap1_z3}\\
  \includegraphics[scale=.47,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/bcrit_mmsn_kap10_z0}\includegraphics[scale=.47,clip=true,trim=2.5cm 0cm 0cm
  1cm]{figures/bcrit_mmsn_kap10_z3} 
  \caption{Dimensionless thermal relaxation timescales $\beta$,
    evaluated at the midplane (left) and at $z=3H$ (right) in the
    fiducial protoplanetary disk. Eq. \ref{beta_mmsn} is plotted  
    for three values of the 
    perturbation radial wavenumber: $\khat=1$ (dotted), $\khat=10$
    (dashed) and $\khat=100$ (dashed-dot), for three values of the
    opacity relative to the MMSN: $\hat{\kappa}_d=0.1$ (top),
    $\hat{\kappa}_d=1$ (middle) and $\hat{\kappa}_d=10$ (bottom).  
    The solid line is the 
    critical thermal relaxation timescale $\beta_\mathrm{crit}$. 
    % and 
    % $\beta<\beta_\mathrm{crit}$ is necessary for the fundamental isothermal VSI to operate  
    % for $\khat\gg1$. 
    \label{mmsn_bcrit_bcool}}   
\end{figure*}

@@

\subsection{Numerical examples}\label{mmsn_example}
Here we solve the linear stability problem with the
thermal relaxation function defined by Eq. \ref{beta_mmsn}. We
consider the MMSN at $r=50$AU and $r=2$AU.

We first compare $\beta$ at these radii in Fig. \ref{beta_compare}. In
the optically thin outer disk, $\beta$ is independent of $z$ and
$\beta<\beta_\mathrm{crit}$ throughout the disk column. We thus expect
the VSI to operate at $r=50AU$. In the inner disk, $\beta$ depends on 
height such that $\beta\lesssim\beta_\mathrm{crit}$ only for 
$|z|\gtrsim 2H$, which suggest that the VSI will be restricted
restricted to the upper and lower layers in the inner disk.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/beta_compare} 
  \caption{Thermal relaxation timescales in the MMSN at $r=50$AU
    (thick solid) and $r=2$AU (thick dotted) for perturbation
    wavenumbers $\khat=10,\,100$, respectively. The corresponding
    horizontal lines are the critical thermal relaxation timescales
    derived in linear theory. 
    \label{beta_compare}}  
\end{figure}

In Fig. \ref{eigen_compare_mmsn} we compute growth rates of the
fundamental mode at $r=50$AU and $r=2$AU. At $r=50$AU growth rates are
insensitive to $\khat$ for $1\lesssim \khat\lesssim 60$, and
corresponds to growth time $\simeq 40$ local orbits. In the inner disk
at $r=2$AU, we only found unstable modes for $\khat\gtrsim50$. To
obtain a similar growth timescale at $r=2$AU requires $\khat\sim
100$. This is consistent with the expectation that, as we proceed into
the inner disk, it bcomes optically thick, so one needs to consider
smaller perturbation wavelengths to obtain a sufficiently short
thermal timescale.

% In both cases we find that increasing $\khat$ increases the
% perturbation amplitudes near the disk surface. Such modes may be
% affected by boundary conditions so their robustness is unclear.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigen_compare_grow_mmsn} 
  \caption{Growth rates of the fundamental VSI mode in the MMSN at
    $r=50$AU (solid) and $r=2$AU (dotted). 
    \label{eigen_compare_mmsn}}  
\end{figure}

Finally, we compare eigenfunctions for the vertical velocity at these
radii in Fig. \ref{mmsn_eigenvz}. We compare cases with similar growth
rates. At $r=50$AU we have vertical motion throughout the disk column
because the thermal timescale is less than critical. By contrast, at
$r=2$AU there is little vertical motion in $|z|\lesssim 2H$. 
Considering this result with Fig. \ref{beta_compare} suggest 
that the critical thermal timescale gives an indication of where the 
VSI operates even when $\beta$ depends on $z$.

%can always go to sufficiently large heights to get beta < beta crit,
%but getting close to boundaries/lower densities

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_50AU}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0cm]{figures/eigenvectorvz_mmsn_2AU}
  \caption{Comparison of vertical velocity perturbations of 
    fundamental VSI modes of similar growth rates in the MMSN at
    $r=50$AU (top) and $r=2$AU (bottom). The real (imaginary) part of
    $\delta v_z$ is shown as 
    the solid (dotted) curves. The eigenfunction has been normalized
    such that $\imag(\delta v_z)=0$ at the vertical boundaries, and
    $\mathrm{max}|\delta v_z|=1$. \label{mmsn_eigenvz}}      
\end{figure}
% only find one unstable mode -> close to marginal 
% larger kx
